
---
title: "Cyclistic Case Study"
output: github_document
date: "2025-02-06"
---

```{r setup, include=FALSE}
library(knitr)

# Automatically save plots in the "plots/" folder
knitr::opts_chunk$set(fig.path = "plots/", fig.width = 10, fig.height = 7.7, echo = TRUE)
```

## Introduction
This capstone project is the final requirement for the Google Data Analytics course, featuring a fictional bike-share company based in Chicago called Cyclistic. The goal of this analysis is to support Cyclist's marketing team in developing a strategy to convert casual riders into annual members. This study aims to identify key differences in how casual riders and annual members use Cyclistic bikes. The data analysis process—Ask, Prepare, Process, Analyze, Share, and Act—will guide this project. The analysis will be conducted using R for data cleaning, exploration, and visualization.

## Ask
Cyclistic aims to understand how annual members and casual riders differ in their usage patterns. Annual members are currently more profitable than casual riders, presenting an opportunity for growth through targeted marketing strategies.

This analysis seeks to:

- Identify usage patterns, trip behaviors, and trends among casual riders and annual members
- Analyze 12 months of historical bike-share data to uncover key insights
- Provide data-driven recommendations to support the marketing team’s goal of increasing annual memberships

The final findings will be presented to the Marketing Director and the Cyclistic Executive Team to inform strategic decisions.

## Prepare

load packages
```{r message=FALSE}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(readr)
library(janitor)
library(geosphere)
library(dplyr)
library(pastecs)
library(scales)
```
<br>
Load data frames. I will be using the 2024 data, which is the most recent data from the last 12 months at the time of this report. 
```{r message=FALSE}
JAN_2024 <- read_csv("Data/202401-divvy-tripdata.csv")
FEB_2024 <- read_csv("Data/202402-divvy-tripdata.csv")
MAR_2024 <- read_csv("Data/202403-divvy-tripdata.csv")
APR_2024 <- read_csv("Data/202404-divvy-tripdata.csv")
MAY_2024 <- read_csv("Data/202405-divvy-tripdata.csv")
JUN_2024 <- read_csv("Data/202406-divvy-tripdata.csv")
JUL_2024 <- read_csv("Data/202407-divvy-tripdata.csv")
AUG_2024 <- read_csv("Data/202408-divvy-tripdata.csv")
SEP_2024 <- read_csv("Data/202409-divvy-tripdata.csv")
NOV_2024 <- read_csv("Data/202411-divvy-tripdata.csv")
OCT_2024 <- read_csv("Data/202410-divvy-tripdata.csv")
DEC_2024 <- read_csv("Data/202412-divvy-tripdata.csv")
```

## Process

Compares columns names/type across multiple data frames. This confirms that the column names and data types are consistent across all data frames used. 
```{r}
compare_df_cols(JAN_2024, FEB_2024, MAR_2024, APR_2024, MAY_2024, JUN_2024, JUL_2024, AUG_2024, SEP_2024, OCT_2024, NOV_2024, DEC_2024)
```
<br>
Merge all 12 data frames for 2024 together to consolidate analysis. 
```{r}
#merge 2024 data frames together
merged_2024 <- bind_rows(JAN_2024, FEB_2024, MAR_2024, APR_2024, MAY_2024, JUN_2024, JUL_2024, AUG_2024, SEP_2024, OCT_2024, NOV_2024, DEC_2024)
```
<br>
Count the number of n/a values in each column to assess data completeness. This results in approximately ~18-19% of missing data from the station names and station id columns. Since this information is not essential for analyzing ride times, ride counts, or seasonality, I have chosen to retain the incomplete data.
```{r}
#count n/a values in column
colSums(is.na(merged_2024))
```
<br>
Add a new column that calculates a trips ride time in minutes by finding the time difference between started_at and ended_at column. I've also added a date
```{r}
merged_2024 <- merged_2024 %>%
  distinct() %>% #remove duplicates
  mutate(ride_time_min = difftime(ended_at, started_at, units = "min")) %>%  #add column that calculates the ride time in minutes
  arrange(desc(ride_time_min)) %>% 
  mutate(day_of_week = wday(started_at, label = TRUE, abbr = FALSE)) %>% #add date_of_week column 
  mutate(ride_month = month(started_at, label = TRUE, abbr = FALSE)) %>% #add ride_month column  %>% 
  mutate(start_time_only = hms(format(started_at, format = "%H:%M:%S")))
```

Remove negative values in ride_time_min column and any ride times greater than 24 hrs. These values are 
```{r}
#remove negative values in ride_time_min column and ride times greater than 24 hrs.
merged_2024 <- merged_2024 %>%  
  filter(ride_time_min > 0, ride_time_min <= 1440)
```

## Analyze
Perform description analysis on ride_time_min column across all users.

- Minimum ride time is 0. This demonstrates that users are quickly un-docking and re-docking bikes without starting a ride. 
- Max ride time is 1439 min. Rides longer than 1440 min (24 hrs) have been removed from the data.
- Median ride time is 9.70 min. Many users are taking shorter trips. 
- Mean is 15.40 which is larger than the medium. The data is right-skewed due to users with exceptionally long ride times. 
```{r}
round(stat.desc(merged_2024$ride_time_min), 2)
```
<br>
Perform descriptive statistical analysis on ride_time_min column by user type.

- Total Rides: Members make up 63.3% of total rides with 3,707,005 rides versus 2,145,244 rides for casual users. 
- Casual riders take longer rides than members with a higher mean (12.02) and median (20.93 min) compared to members.
- Casual riders have a higher standard deviation (42.46 min), meaning their ride times vary widely.
```{r}
# Descriptive statistics for members
round(stat.desc(merged_2024$ride_time_min[merged_2024$member_casual == "member"]),2)

# Descriptive statistics for casual users
round(stat.desc(merged_2024$ride_time_min[merged_2024$member_casual == "casual"]), 2)
```

Calculate the average ride time by month and user type. 
```{r}
#calculate avg ride_time by month and user type
ride_time_month <- merged_2024 %>% 
  group_by(member_casual, ride_month) %>% 
  summarise(mean = mean(ride_time_min))

as_tibble(ride_time_month)
```

Calculate the number of rides by month and user type. 
```{r}
num_rides_month <- merged_2024 %>% 
  count(member_casual, ride_month, name = "num_rides")

as_tibble(num_rides_month)
```

Find the top 10 start stations for member and casual users. 
```{r}
top_stations <- merged_2024 %>% 
  count(member_casual, start_station_name, name = "station_count") %>% 
  arrange(desc(station_count)) %>% 
  filter(!is.na(start_station_name)) %>% 
  group_by(member_casual) %>% 
  top_n(10, station_count)

as_tibble(top_stations)
```

Find the count of bike usage based on bike type between casual and member riders.
```{r}
bike_type <- merged_2024 %>% 
  group_by(member_casual, rideable_type) %>% 
  summarize(count = n())

as_tibble(bike_type)
```
Calculate ride counts by day of the week among all users as well as between user type. 
```{r}
#all users
table(merged_2024$day_of_week)

#member vs. casual
day_count <- merged_2024 %>% 
    group_by(member_casual, day_of_week) %>% 
    summarize(ride_count = n())

as_tibble(day_count)
```

Calculate number of ride by month between user type.
```{r}
#ride count by month
month_count <- merged_2024 %>% 
  group_by(member_casual, ride_month) %>% 
  summarize(ride_count = n())

as_tibble(month_count)
```

## Share

#### Ride Duration
For members, the higher density peak at ~4 minutes suggests they frequently take short, consistent rides. In contrast, casual users have a lower density peak at ~7 minutes, indicating longer rides on average but with greater variability in ride duration. Both groups exhibit a right-skewed distribution, with casual users' distribution being slightly higher in the right tail. This suggests that while longer rides occur in both groups, a higher proportion of casual users take extended rides. 
```{r}
quantile(merged_2024$ride_time_min, probs = c(0.95, 0.99, 0.999), na.rm = TRUE)

merged_2024_filtered_ride_time <- merged_2024 %>%
  filter(ride_time_min <= 42.2)

ggplot(merged_2024_filtered_ride_time, aes(ride_time_min, color=member_casual)) + 
  geom_density() +
  xlim(0, 42.2) +
  theme_minimal() +
  labs(title = "Ride Duration Density by Member and Casual Users",
       x = "Ride Duration",
       y = "Density",
      color = "User Type")
```

On a month to month basis, casual users on average take longer rides than members. This suggests that casual riders are using the bikes for leisure and are more likely to opt for full-day passes. The months with the highest ride duration for both casual users and members are the summer months May through July. The winter months Nov through Jan have the lowest ride duration. For members, ride duration increases less dramatically during peak months compared to casual riders, indicating more consistent usage throughout the year.

```{r}
#plots the average ride duration by month between casual and members
ggplot(ride_time_month, aes(x = ride_month, y = mean, color = member_casual, group = member_casual)) +
  geom_line() + 
  labs(title = "Average Ride Duration by Month",
       x = "Month",
       y = "Average Ride Duration (minutes)",
       color = "User Type") +
  theme_minimal() +
  scale_y_continuous() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
#### Seasonality
Although members are taking shorter rides, they are taking more rides than casual users. The number of rides increases for all users from June to September, followed by a sharp decline from November to January. For both groups, this shows that seasonality makes an impact on number of rides for both groups. 
```{r}
  ggplot(month_count, aes(x = ride_month, y = ride_count, , color = member_casual, group = member_casual)) + 
    geom_line() +
    scale_y_continuous(labels = comma) +
    labs(title = "Ride Count By Month and User Type - 2024", 
         x = "Day of the Week", 
         y = "Ride Count",
         fill = "User Type") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


#### Weekday vs. Weekend Ridership
On a weekly basis, members take more rides on the weekday and casual members take more rides on the weekends. For members, Wednesday has the highest ride count, with Saturday being the lowest. For casual user, we see the opposite. Saturday had the highest ride count with Tuesday having the lowest. 
```{r}
  ggplot(day_count, aes(x = day_of_week, y = ride_count, , color = member_casual, group = member_casual)) + 
    geom_line() +
    scale_y_continuous(labels = comma) +
    labs(title = "Ride Count By Day of the Week and User Type - 2024", 
         x = "Day of the Week", 
         y = "Ride Count",
         fill = "User Type") +
    theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Members tend to start their rides between 8:00 AM and 10:00 AM before experiencing a slight dip. Around noon, ride activity begins to increase again, peaking at approximately 5:00 PM. In contrast, casual riders do not exhibit noticeable spikes in the morning. Instead, their ride counts steadily rise from 8:00 AM, reaching a peak around 5:00 PM. The busiest period for casual riders is between 2:00 PM and 7:00 PM.
```{r}
ggplot(merged_2024, aes(x = start_time_only, fill = member_casual)) +
  geom_histogram(binwidth = 15 * 60, alpha = 0.4, position = "identity") +
  scale_x_time(
    breaks = seq(0, 24 * 60 * 60, by = 2 * 60 * 60),  # 2-hour breaks (in seconds)
    labels = scales::time_format("%H:%M")
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Ride Start Time Distribution",
    x = "Time of Day",
    y = "Ride Count",
    fill = "User Type"
  ) +
  theme_minimal()
```

#### Top stations
The top 10 start stations for casual members show they are starting their rides around popular tourist areas. 
```{r}
#Bar chart for top start stations between casual and member riders
ggplot(top_stations, aes(x = reorder(start_station_name, station_count), y = station_count, fill = member_casual)) +
  geom_col() +
  coord_flip() +
  facet_wrap(~ member_casual, scales = "free_y") +
  labs(title = "Top 20 Stations by User Type - 2024", 
       x = "Station Name", 
       y = "Ride Count",
       fill="User Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Bike Type
The most used bike types for all members are the electric bikes followed by classic bikes. Electric scooters are the least popular, but there will need to be further investigation to see if this is due to availability of bike types. 
```{r}
#Bar chart of bike type usage between users
bike_type_plot <- ggplot(bike_type, aes(x = rideable_type, y = count, fill = member_casual)) +
  geom_col() +
  scale_y_continuous(labels = comma) +
  facet_wrap(~member_casual) + 
  labs(title = "Bike Type Usage Between User Types - 2024", 
       x = "Bike Type", 
       y = "Ride Count",
       fill = "User Type")+
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#ggsave("plots/bike_type_usage.png", plot = bike_type_plot, width = 6, height = 4, dpi = 300)

bike_type_plot
```

## Insight summary and recommendations

### 1. Seasonality Drives Demand:

#### Summary: 

- Both casual and member riders increase usage in the summer (May–July) and drop significantly in winter (Nov–Jan).
- Casual riders show more extreme seasonal fluctuations, while members maintain a more consistent level of use.

#### Recommendation: 

- Offer discounted rates for memberships during off-peak seasons
- Provide tips and messaging on how to prepare for rides in the winter months.

###  2. Casual Riders Take Longer Trips:

#### Summary: 

- Casual riders have a higher average ride time (20.93 min vs. 12.02 min for members) and a higher standard deviation in ride duration.
- This suggests that casual riders are using bikes more for leisure, sightseeing, or long trips, while members likely use bikes for commuting or short, frequent trips.

#### Recommendation:

- Offer discounted full-day membership packs for frequent casual riders. For example, 10-pack of full day rides. 
- Consider partnerships with local tour groups or create a multi-day sightseeing guide to complement full-day ride plans. 

### 3. Members Ride More Often, Especially on Weekdays:

#### Summary:

- Members account for 63.3% of total rides despite taking shorter trips.
- Their rides peak on weekdays, especially Wednesdays, suggesting they use the service for commuting.
- Casual riders peak on weekends, indicating more recreational use.

#### Recommendation:

- Introduce a weekend only membership plan at a discounted rate.




